### 2016-03-23 mcmc.tex
words about convergence and tuning

\begin{problem}\label{prob:twod}
Re-do \problemname~\ref{prob:MH} but now with an input density
that is a function of two variables $(x, y)$.
For the density function use two different functions.
The first density function is a covariant two-dimensional Gaussian density with variance tensor
\begin{eqnarray}
V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
\quad .
\end{eqnarray}
The second density function is a rectangular top-hat function that is uniform on
the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
density with zero mean and variance tensor set to the two-dimensional
identity matrix.
Plot the two one-dimensional histograms and also a two-dimensional scatter
plot for each sampling.
\end{problem}

@@

The autocorrelation function is defined by
\begin{eqnarray}
C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
\label{eq:autocorrelationfn}\\
\mu &=& E[\pars]
\quad ;
\end{eqnarray}
that is, it is the covariance of points separated by $\Delta$, taken
relative to the variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau = 1 + 2\,\sum_{\Delta = 1}^{\infty} C(\Delta)
\label{eq:autocorrelationtime}\quad .
\end{eqnarray}
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

That is, if the expectation value $E[q]$ is computed by the sampling
approximation given in \equationname~(\ref{eq:expectationvalue}),
there is an uncertainty on that expectation given by
\begin{eqnarray}
  \Var[E[q]] &=& \frac{\tau}{K}\,\Var[q]
  \\
  &\approx& \frac{\tau}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) - E[q]\right)^2
  \quad ,
\end{eqnarray}
where we are being a bit loose with terminology because this is the
uncertainty not on $E[q]$ but on our sampling-based \emph{estimator} for
$E[q]$, and the $E[q]$ inside the sum is that estimator too.
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

@@

Most \MCMC\ methods
  (though not all\note{We are promoters of an ensemble method that has a
  nearly-fixed proposal distribution and requires no tuning \citep{emcee}.
  There are many problems for which this is \emph{not} the best choice
  of sampler, but since it obviates tuning, it can be a good first choice.})
  make use of something like a ``proposal distribution''
  which determines what kinds of steps the walker can take
  as it random-walks through the parameter space.
The user generally has a lot of control over what this proposal distribution might be,
  and how to choose the parameters.
For example, in a $D$-dimensional parameter space, %do we consistently use d here?
  a zero-mean but anisotropic Gaussian (normal distribution)
  is often used to draw offsets for the walker to take from point to point.
Even within this choice (which is of one function among many possibilities),
  there are $D\,(D+1)/2$ parameters in the $D\times D$ symmetric, positive definite covariance matrix
  to set ``by hand''.
How to choose this function and set these parameters?

@@

When sampling really gets very slow,
  there are various tricks and tips to work through.
One is to see if any of the parameters has a conditional posterior
  probability that is guaranteed to be Gaussian
  (this is true for some linear parameters in least-square problems
  when the priors are Gaussian, for example).
These dimensions can often be sampled \emph{exactly},
  outside of the MCCM framework;
  the algorithm looks like:
Take an \MCMC\ step in the hard parameters,
  then take an exact step in the easy parameters,
  repeat.
Another related trick is to analytically marginalize,
  rather than sample,
  easy nuisance parameters.
The huge bag of possible tricks is so large,
  it goes way beyond the scope of this introductory \documentname.
In our experience, it is valuable to make friends with at least one statistician,
  one applied mathematician, and one computer scientist.
Between them, they ought to span the literature.

>>>>

\begin{problem}\label{prob:twod}
Re-do \problemname~\ref{prob:MH} but now with an input density
that is a function of two variables $(x, y)$.
For the density function use two different functions.
The first density function is a covariant two-dimensional Gaussian density with variance tensor
\begin{eqnarray}
V &=& \left[\begin{array}{cc} 2.0 & 1.2 \\ 1.2 & 2.0 \end{array}\right]
\quad.
\end{eqnarray}
The second density function is a rectangular top-hat function that is uniform on
the joint constraint $3<x<7$ and $1<y<9$ and zero everywhere else.
For the proposal distribution $q(x', y'\given x, y)$ a two-dimensional Gaussian
density with zero mean and variance tensor set to the two-dimensional
identity matrix.
Plot the two one-dimensional histograms and also a two-dimensional scatter
plot for each sampling.
\end{problem}

@@

The autocorrelation function is defined by
\begin{eqnarray}
C_f(\Delta) &\equiv& \frac{E[(f(\pars_k)-\mu_f)\,(f(\pars_{k+\Delta})-\mu_f)]}
                          {E[(f(\pars)-\mu_f)^2]}
\label{eq:autocorrelationfn}\\
\mu_f &=& E[f(\pars)]
\quad ;
\end{eqnarray}
where $f(\pars)$ is a scalar valued function of the parameters.
In other words, \equationname~(\ref{eq:autocorrelationfn}) is the covariance
of points in the chain separated by $\Delta$ steps, taken relative to the
variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C_f(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C_f(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau_f = 1 + 2\,\sum_{\Delta = 1}^{\infty} C_f(\Delta)
\label{eq:autocorrelationtime}\quad.
\end{eqnarray}
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

That is, if the expectation value $E[q]$ is computed by the sampling
approximation given in \equationname~(\ref{eq:expectationvalue}),
there is an uncertainty on that expectation given by
\begin{eqnarray}
  \Var[E_{p(\pars)}[q(\pars)]] &=& \frac{\tau_q}{K}\,\Var_{p(\pars)}[q(\pars)]
  \\
  &\approx& \frac{\tau_q}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) -
    E_{p(\pars)}[q(\pars)]\right)^2
  \quad,
\label{eq:tauvar}
\end{eqnarray}
where we are being a bit loose with terminology because this is the
variance not on $E_{p(\pars)}[q(\pars)]$ but on our sampling-based
\emph{estimator} for
$E_{p(\pars)}[q(\pars)]$, and the $E_{p(\pars)}[q(\pars)]$ inside the sum is
that estimator too.
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

It is also worth noting that there is not just one autocorrelation function or
autocorrelation time for an MCMC chain.
Instead, there is an autocorrelation time $\tau_f$ for every function
$f(\pars)$ and these will, in general, be different.
Therefore, the autocorrelation time used in \equationname~(\ref{eq:tauvar}) is
the time computed on the chain of $q(\pars)$ values instead of the
autocorrelation time for $\pars$.

@@

Most \MCMC\ methods make use of something like a ``proposal distribution''
  which determines what kinds of steps the walker can take
  as it random-walks through the parameter space.
The user generally has a lot of control over what this proposal distribution might be,
  and how to choose the parameters.
For example, in a $D$-dimensional parameter space, %do we consistently use d here?
  a zero-mean but anisotropic Gaussian (normal distribution)
  is often used to draw offsets for the walker to take from point to point.
Even within this choice (which is of one function among many possibilities),
  there are $D\,(D+1)/2$ parameters in the $D\times D$ symmetric, positive definite covariance matrix
  to set ``by hand''.
How to choose this function and set these parameters?

@@

When sampling really gets very slow,
  there are various tricks and tips to work through.
% One is to see if any of the parameters has a conditional posterior
%   probability that is guaranteed to be Gaussian
%   (this is true for some linear parameters in least-square problems
%   when the priors are Gaussian, for example).
% These dimensions can often be sampled \emph{exactly},
%   outside of the \MCMC\ framework;
%   the algorithm looks like:
% Take an \MCMC\ step in the hard parameters,
%   then take an exact step in the easy parameters,
%   repeat.
% Another related trick is to analytically marginalize,
%   rather than sample,
%   easy nuisance parameters.
The huge bag of possible tricks is so large,
  it goes way beyond the scope of this introductory \documentname.
In our experience, it is valuable to make friends with at least one
  statistician, one applied mathematician, and one computer scientist.
Between them, they ought to span the relevant literature.
